Estimating Body Fat Percentage Using Anthropometric Measurements

### 2024-12-03 paper/paper.qmd
updated some formating
made some lintr changes & changes some mistakes

---
title: "Estimating Body Fat Percentage"
subtitle: "Abdomen as the Strongest Predictor With Contributions From Wrist, Height, and Age For Affordable Health Monitoring"
author: 
  - Wendy Yuan
thanks: "Code and data are available at: https://github.com/kiwindyy/Body-Fat"
date: November 29, 2024
date-format: long
abstract: "This study uses a multi-linear regression model to estimate body fat percentage based on circumference measurements. Abdominal size was the strongest predictor, followed by wrist size, height, and age, which showed minimal contributions. These results demonstrate that accessible methods can effectively estimate body fat, offering methods suitable for non-clinical settings. This approach is important for identifying and managing health risks associated with extremely low or high body fat levels in a practical and affordable way."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

Body fat percent is measured in percentage (%). Using the method from ... body fat percentage is calculated using the relationship between body density ($D$) and the densities of lean and fat tissue. The formula for the proportion of fat tissue ($B$) is:

@@

- Density: The measure of body density, used as a key variable in estimating body fat percentage. It combines body weight and volume to provide insights into overall composition. It is measured in grams per $cm^3$ with values ranging from 0.99 to 1.1.
- Age: The age of the individual in years, providing a measure of the person's stage of life, which can influence body composition. It is measured in years with values ranging from 20 to 81.
- Weight: The total body mass of the individual, reflecting the combined weight of bones, muscles, fat, and other tissues. It is measured in pounds (lbs) with values ranging from 118.5-262.75.
- Height: The standing height of the individual, often used alongside weight to calculate proportions and indices. It is measured in inches (in) with values ranging from 64–77.75.
- Neck: The circumference of the neck, offering a measurement of fat and muscle distribution in the upper body. It is measured in centimeters (cm) with values ranging from 31.1-43.9.
- Chest: The circumference of the chest, reflecting the size and structure of the upper torso, including muscle and fat. It is measured in centimeters (cm) with values ranging from 79.3-128.3.
- Abdomen: The circumference around the abdomen, a key indicator of central fat distribution and a significant predictor of body fat percentage. It is measured in centimeters (cm) with values ranging from 69.4-126.2.
- Waist: The measurement around the waist, often used with hip circumference to assess body shape and fat distribution. It is measured in centimeters (cm) with values ranging from 27.3-49.6.
- Hip: The circumference of the hips, providing a measure of lower-body proportions and often paired with waist circumference to calculate ratios. It is measured in centimeters (cm) with values ranging from 85.0-125.6. 
- Thigh: The circumference of the thigh, measured to understand muscle and fat distribution in the upper leg. It is measured in centimeters (cm) with values ranging from 47.2-74.4.
- Knee: The circumference of the knee, offering additional detail about lower-body structure and proportions. It is measured in centimeters (cm) with values ranging from 33.5-46.0.
- Ankle: The measurement around the ankle, reflecting skeletal and soft tissue composition in the lower leg. It is measured in centimeters (cm) with values ranging from 20.2-27.0.
- Bicep: The circumference of the bicep, measured with the arm flexed, highlighting upper-arm muscle and fat distribution. It is measured in centimeters (cm) with values ranging from 25.6-39.1.
- Forearm: The circumference of the forearm, providing information about the composition of the lower arm. It is measured in centimeters (cm) with values ranging from 24.6-33.8.
- Wrist: The circumference of the wrist, a useful proxy for skeletal size and overall body frame. It is measured in centimeters (cm) with values ranging from 15.8-21.4.

@@

@@

@@

The regression results are summarized in [@tbl-multi1], showing the coefficients and pvalues & [@tbl-multi2], showing the $R^2$ and standard error. The intercept, which represents the estimated body fat percentage when all predictors are zero, is 2.01. However, this value doesn't have much real-world meaning because variables like height and abdomen size cannot actually be zero. Among the predictors, Abdomen circumference has the strongest positive effect on body fat percentage. For every 1 cm increase in abdomen size, body fat percentage increases by about 0.78%, while keeping other factors unchanged.

@@

Future research should focus on expanding the dataset to include more diverse groups. Adding data for women and people from different backgrounds would make the model more useful for a wider population. Including people of all ages, especially teenagers and seniors, would help the model predict body fat for everyone, not just middle-aged adults. A more balanced dataset would also make it easier to detect patterns that apply to specific groups.

Researchers should also consider adding more predictors to the model. Including variables like weight, hip size, and thigh size could give a fuller picture of body fat distribution and improve the model’s accuracy. Looking at how important each variable is could help decide which ones to add while keeping the model manageable.

Finally, using different methods to model the data could address some of the limitations. For example, machine learning techniques or non-linear models could find patterns that linear regression might miss, especially for outliers. Future studies could also use data collected over time to see how body fat changes as people age and whether the same predictors still apply. This would help create a better understanding of body fat and how it changes throughout life.

@@

***Overview of Related Methods***

@@

***Strengths of the Study***

@@

***Limitations of the Study***

@@

This paper builds on @potter work by addressing some of the identified shortcomings. For example, while @potter focused solely on abdominal circumference, this study incorporates additional variables, such as wrist size and height, into a statistical model. These extra measurements aim to make predictions more accurate, especially for individuals at the lower or higher ends of the body fat range.

Additionally, @potter highlighted that body fat is distributed differently between genders. This study focuses only on middle-aged men. By concentrating on this group, the research aims to provide a method that better captures the specific fat distribution patterns and characteristics of this demographic.

***Implications for Practical Applications***

>>>>

---
title: "Estimating Body Fat Percentage Using Anthropometric Measurements"
subtitle: "Abdomen as the Strongest Predictor With Contributions From Wrist, Height, and Age For Affordable Health Monitoring"
author: 
  - Wendy Yuan
thanks: "Code and data are available at: https://github.com/kiwindyy/Body-Fat"
date: November 29, 2024
date-format: long
abstract: "This study uses a multi-linear regression model to estimate body fat percentage based on circumference measurements. Abdominal size was the strongest predictor, followed by wrist size, height, and age, which showed minimal contributions. These results demonstrate that accessible methods can effectively estimate body fat, offering methods suitable for non-clinical settings. This approach is important for identifying and managing health risks associated with extremely low or high body fat levels in a practical and affordable way."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

Body fat percent is measured in percentage (%). The method and numbers (for the example) are from @katch. Body fat percentage is calculated using the relationship between body density ($D$) and the densities of lean and fat tissue. The formula for the proportion of fat tissue ($B$) is:

@@

- Density: The measure of body density, used as a key variable in estimating body fat percentage. It combines body weight and volume to provide insights into overall composition. It is measured in grams per $cm^3$ with values ranging from 0.99 to 1.1.
- Age: The age of the individual in years, providing a measure of the person's stage of life, which can influence body composition. It is measured in years with values ranging from 22 to 81.
- Weight: The total body mass of the individual, reflecting the combined weight of bones, muscles, fat, and other tissues. It is measured in pounds (lbs) with values ranging from 118.5-262.75.
- Height: The standing height of the individual, often used alongside weight to calculate proportions and indices. It is measured in inches (in) with values ranging from 64–77.75.
- Neck: The circumference of the neck, offering a measurement of fat and muscle distribution in the upper body. It is measured in centimeters (cm) with values ranging from 31.1-43.9.
- Chest: The circumference of the chest, reflecting the size and structure of the upper torso, including muscle and fat. It is measured in centimeters (cm) with values ranging from 79.3-128.3.
- Abdomen: The circumference around the abdomen, a key indicator of central fat distribution and a significant predictor of body fat percentage. It is measured in centimeters (cm) with values ranging from 69.4-126.2.
- Waist: The measurement around the waist, often used with hip circumference to assess body shape and fat distribution. It is measured in inches (in) with values ranging from 27.3-49.6.
- Hip: The circumference of the hips, providing a measure of lower-body proportions and often paired with waist circumference to calculate ratios. It is measured in centimeters (cm) with values ranging from 85.0-125.6. 
- Thigh: The circumference of the thigh, measured to understand muscle and fat distribution in the upper leg. It is measured in centimeters (cm) with values ranging from 47.2-74.4.
- Knee: The circumference of the knee, offering additional detail about lower-body structure and proportions. It is measured in centimeters (cm) with values ranging from 33.0-46.0.
- Ankle: The measurement around the ankle, reflecting skeletal and soft tissue composition in the lower leg. It is measured in centimeters (cm) with values ranging from 19.1–33.9.
- Bicep: The circumference of the bicep, measured with the arm flexed, highlighting upper-arm muscle and fat distribution. It is measured in centimeters (cm) with values ranging from 25.6-39.1.
- Forearm: The circumference of the forearm, providing information about the composition of the lower arm. It is measured in centimeters (cm) with values ranging from 21.0–34.9.
- Wrist: The circumference of the wrist, a useful proxy for skeletal size and overall body frame. It is measured in centimeters (cm) with values ranging from 15.8-21.4.

@@

par(cex = 0.8)

@@

The multi-linear regression model was chosen for its simplicity, ease of understanding, and ability to show important relationships between measurements while meeting the basic requirements of linear regression, such as consistency and fairness in the results. Unlike more complicated methods, such as machine learning or non-linear models, multi-linear regression makes it easier to interpret how each factor affects body fat estimation, making it more useful for practical applications and understandable for non-experts. This model connects the findings with existing knowledge by confirming known patterns, such as the strong connection between abdominal size and body fat, which aligns with the idea that fat around the abdomen is a main indicator of overall fat levels. Including wrist size and height emphasizes differences in body structure and proportions, adding more detail to how body fat is distributed. By combining known ideas with real-world testing, this model provides a practical, clear, and easy way to estimate body fat, especially for everyday use, while building on current understanding.

@@

The regression results are summarized in [@tbl-multi1], showing the coefficients and pvalues & [@tbl-multi2], showing the $R^2$ and standard error. For the code and full model see the script document in: scripts/06-model_data.R. The intercept, which represents the estimated body fat percentage when all predictors are zero, is 2.01. However, this value doesn't have much real-world meaning because variables like height and abdomen size cannot actually be zero. Among the predictors, Abdomen circumference has the strongest positive effect on body fat percentage. For every 1 cm increase in abdomen size, body fat percentage increases by about 0.78%, while keeping other factors unchanged.

@@

Future research should focus on expanding the dataset to include more diverse groups. Adding data for women and people from different backgrounds would make the model more useful for a wider population. Including people of all ages, especially teenagers and seniors, would help the model predict body fat for everyone, not just middle-aged adults. A more balanced dataset would also make it easier to detect patterns that apply to specific groups. Researchers should also consider adding more predictors to the model. Including variables like weight, hip size, and thigh size could give a fuller picture of body fat distribution and improve the model’s accuracy. Looking at how important each variable is could help decide which ones to add while keeping the model manageable.

Finally, using different methods to model the data could address some of the limitations. For example, machine learning techniques or non-linear models could find patterns that linear regression might miss, especially for outliers. Future studies could also use data collected over time to see how body fat changes as people age and whether the same predictors still apply. This would help create a better understanding of body fat and how it changes.

@@

@@

***Strengths & Limitations of the Study***

@@

@@

This paper builds on @potter work by addressing some of the identified shortcomings. For example, while @potter focused solely on abdominal circumference, this study incorporates additional variables, such as wrist size and height, into a statistical model. These extra measurements aim to make predictions more accurate, especially for individuals at the lower or higher ends of the body fat range. Additionally, @potter highlighted that body fat is distributed differently between genders. This study focuses only on middle-aged men. By concentrating on this group, the research aims to provide a method that better captures the specific fat distribution patterns and characteristics of this demographic.
